---
output: 
  github_document:
    toc: true
    toc_depth: 4
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# FlowSoFine

<!-- badges: start -->
<!-- badges: end -->

The goal of FlowSoFine is to quickly and easily generate data structures that make it easy to get statistical output starting from raw fcs-files and a metadata table. The grid approach is especially useful if there aren't any populations clearly visible in the files. This works with a flexible amount of channels and correction for spatial dependencies.


## Installation

You can install the development version of FlowSoFine from GitHub with:

``` r
library(devtools)
install_github("JonasKup/FlowSoFine")
```

## Quickstart


This is a basic example which shows you how to solve a common problem:

```{r example, warning = FALSE}
library(FlowSoFine)
setwd("C:/Users/jonas/Documents/IUF/FACSDATA/testExp/fcs")


#for reading fcs files. Installation via Bioconductor.
library(flowCore)



#for community ecology statistics
library(vegan)



fcs <- read.flowSet(pattern = ".fcs")
metadata <- read.csv2("metadata.csv")

#Create the CoreTemplate
ct <- CoreTemplate(flowset = fcs, channels = c("FSC PAR", "Hoechst Red"), resolution = 50)

#plot the first sample of the CoreTemplate
plot(template = ct, sample = 1)


#Create a distance matrix with adjustments for spatial dependencies
distM <- weightedBray(ct)

#You can use the adonis function from the vegan package to get statistical output
adonis2(distM ~ treatment + age, data = metadata)

```

## Multidimensional analysis
