---
output: 
  github_document:
    toc: true
    toc_depth: 4
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# FlowSoFine

<!-- badges: start -->
<!-- badges: end -->

The goal of FlowSoFine is to quickly and easily generate data structures that make it easy to get statistical output starting from raw fcs-files and a metadata table. The grid approach is especially useful if there aren't any populations clearly visible in the files. This works with a flexible number of channels and correction for spatial dependencies.


## Installation

You can install the development version of FlowSoFine from GitHub with:

``` r
library(devtools)
install_github("JonasKup/FlowSoFine")
```

## Quickstart


This is a basic example which shows you how to solve a common problem:

```{r example}
library(FlowSoFine)
setwd("C:/Users/jonas/Documents/IUF/FACSDATA/testExp/fcs")


#for reading fcs files. Installation via Bioconductor. https://www.bioconductor.org/packages/release/bioc/html/flowCore.html
library(flowCore)

fcs <- read.flowSet(pattern = ".fcs")

```

```{r exhidden, echo = FALSE}
sampleNames(fcs) <- paste("sample", sep = "_", 1:length(sampleNames(fcs)))
```

```{r example2}

metadata <- read.csv2("metadata.csv")

#Create the FSFTemplate
template <- FSFTemplate(flowset = fcs, channels = c("FSC PAR", "Hoechst Red"), resolution = 35)

#plot the first sample of the FSFTemplate
plot(template = template, sample = 1)


#Create a distance matrix with adjustments for spatial dependencies
distM <- weightedBray(template)


#for community ecology statistics
library(vegan)

#You can use the adonis2 function from the vegan package to get statistical output
adonis2(distM ~ treatment + age, data = metadata)

```

## Plotting
```{r Plotting}
#plot a flowCore flowFrame with a flexible resolution and transformation function
plotFF(flowFrame = fcs[[1]], channels = c("FSC PAR", "Hoechst Red"), resolution = 250, transformation = asinh)


#plot the first sample of the above created FSFTemplate
plot(template, sample = 1)

#add info to the plot using the ggplot2 library
library(ggplot2)
#add tbe total number of events in each bin for sample 1 to the plot
plot(template, sample = 1) + geom_text(aes(label = template@counts[,1]), size = 2)

#add the bin number to each bin
plot(template, sample = 1) + geom_text(aes(label = 1:template@nBins), size = 2)


```

## T-Scores

## Additional statistical output (adonis2, NMDS)

## Multidimensional analysis

### Creating and converting higher dimensional FSFTemplates

### Plotting with less or more than two channels

### Analysing single channels
